---
title: "Applied Bayesian Analyses with R (Geneva - 2024)"
output: 
  flexdashboard::flex_dashboard:
    orientation: columns
    vertical_layout: scroll
---


```{r setup, include=FALSE}
library(flexdashboard)
```

# Intro {.sidebar}

This site bundles all the material used for the course [ABAR_Geneva24]{style="color: #447099"}. 

---


Instructor: 

 - <a href="https://sdemaeyer.quarto.pub/" target="_blank">Sven De Maeyer</a>
 

---

# Quick Overview

## Column 1

### Outline

In a broad range of scientific disciplines, Bayesian statistics are gaining popularity. Yet the basic training of most researchers only introduces the frequentist framework for statistical inference. This workshop aims to introduce Bayesian statistics in a practical way, laying a foundation for crucial concepts in the Bayesian realm.
 
Participants will be introduced to a workflow for Bayesian analyses in the open-source environment `R`. The starting point will be the linear model (aka regression model) and we can extend this to linear mixed models. The central piece of software will be the `brms` package in `R`, that bridges the typical R modelling language with `Stan`, which is a probabilistic programming language built to estimate models within the Bayesian framework. Given that a workflow in `R` will be introduced, it is advised that participants have some experience working with `R`. Moreover, packages will be used from the tidyverse (e.g., `dplyr`, `ggplot2`, …). Therefore, we advise participants to acknowledge themselves with these packages. 

## Column 2

### Rooms
| When?  | Where    |
|:--------|:-------|
| Tuesday | H4-02-A|
| Wednesday | H8-01-F|
| Thursday | H4-02-A|
| Friday | H4-02-A|

# Preparation

## Column 1

### Intro

For the workshop we will be using R. To be able to participate properly, it is advisable to install a number of packages on your own PC beforehand, as well as to become familiar with the way I usually code in R. This document is a brief guide to help you prepare.

### Necessary Software & Packages

#### CmdStan & CmdStanR

Bayesian estimation and analyses require some dedicated samplers that should be installed on your machine. The workflow that we will cover in the workshop relies on `Stan` software. `Stan` stands alone from R but can be called through R making use of two different ways. We will cover this more in the workshop. 

For the workshop I will use the `CmdStan` chain and the package `cmdstanr`. Therefore I would encourage you to download and install a working `cmdstan` and `cmdstanr`. The following vignette can help you accomplish this: https://mc-stan.org/cmdstanr/articles/cmdstanr.html 

Normally if you follow the steps described in that vignette you will succeed in installing and testing your installation. 

#### `brms`

For the analyses we will rely on the amazing package `brms`. Installing this package is not that difficult, but maybe install before the workshop and start some exploring ;-)

#### `tidyverse`

When I code I use the <b> functional programming </b> approach and a lot of the <b>tidy</b> principles (see below). The package `tidyverse` bundles a set of packages that are needed for this approach to coding. If you want to learn more on the whole universe of `tidyverse` you can explore the book of Hadley Wickham: https://r4ds.had.co.nz/


#### `ggplot2`

Part of the `tidyverse` package is the `ggplot2` package. Therefore it will be installed if you also install `tidyverse` normally. Nevertheless, I find it worth mentioning separately because I think knowing how to make graphs with `ggplot2` is very handy even if you do not dive into bayesian analyses. A great resource to learn to visualize your data is the tutorial written by Cédric Scherer: https://www.cedricscherer.com/2019/08/05/a-ggplot2-tutorial-for-beautiful-plotting-in-r/ !

He also blogs on creating great visualizations and has some nice talks that I think you can find on YouTube!

#### `patchwork`

I mention this package separately as well. `patchwork` allows you to combine different plots created with `ggplot2`. This package should be installed on it's own.

#### `tidybayes`

This package is a dedicated package to make use of the <b>tidy</b> principles when applying it to (results of) Bayesian analyses.

#### `bayesplot`

We will use visualisations a lot in Bayesian analyses to summarize the information of models and parameter estimation. `bayesplot` bundles a number of functions that make this visualizing much easier (although it uses `ggplot2` under the hood so we could accomplish similar results just knowing `ggplot2`).

### Coding in the tidy universe...

There are a great number of roads that will lead to Rome! Coding in R can be accomplished in several ways. This means that code can get messy as well! Therefore, I try to keep myself as much as possible to some rules that are described by what is called the <b>tidy</b> way of coding. Some basic things are key to this approach. I will shortly describe them here so you will recognize this way of working when we're in the workshop. 

#### The pipe

When coding I make use of what is called <b>the pipe operator</b> (notice that you can only use this when you loaded the `tidyverse` package).

The pipe in R code is the following: ```%>%```

What is nice about using the pipe operator is that you can read a piece of code from left to right focussing on the verbs (as Hadley Wickham will call it sometimes).

A short example. 

Imagine you have a vector of numbers for which you want to calculate a mean: ```c(1,2,3,4,5)```.

There are different ways to do this in R. 

A first one is doing it in separate lines of code:

```{r}
x <- c(1,2,3,4,5)
mean(x)
```

For this easy example this would work fine. But a typical side-effect of working this way is that we create a lot of objects along the way. Here we created the object `x` that stored the values of the vector.

To avoid this we could combine all the code in one line by embedding functions:

```{r}
mean(c(1,2,3,4,5))
```

One line of code and no storage of a new object. But this can get complicated if you want to combine a set of functions resulting in almost non-readable code! 

Then there is the pipe operator:

```{r}
library(tidyverse)
c(1,2,3,4,5) %>%
  mean()
```

This is how to read this code:

```{r}
c(1,2,3,4,5) %>% # create the vector and then
  mean()         # calculate the mean
```

So the pipe can be read as `take (or create) A` and `then do B` with the result. We could create a whole pipeline of functions to be applied in this way making use of the pipe operator (this becomes more clear in the next section when we shortly describe the `dplyr` verbs).

A tweet that I encountered gives a great analogy. Using the pipe you could write:

> I %>% woke up %>% took a shower %>% got breakfast %>% took the metro %>% arrived at work %>% ...

In a recent update of `R` they also introduced this idea of a pipe as a "native pipe". This is written as `|>` and functions in a similar way as `%>%` with the advantage that it will also work outside the `tidyverse`. I'm slowly transforming all my material and code to make use of this native pipe. But as I am notoriously sloppy, the chances are high that you will encounter some old-fashioned tidyverse pipes in my code during the workshop...

> I |> learned about the native pipe |> transformed my code |> potentially failed somewhere ...

#### `dplyr` verbs

Within the `tidyverse` universe we can make use of a great package called `dplyr` for a broad spectrum of data-management functions. The functions provided in that package are also sometimes called <b>verbs</b>:

- `mutate( )` to do recoding etc;
- `select( )` to select a subset of columns/variables;
- `filter( )` to filter cases;
- `group_by( )` to apply what's coming afterwards for specific groups in the data by a grouping variable;
- `arrange( )` to sort the data;
- `summarize( )` to apply summarizing functions (like taking the mean etc.);
- `rename( )` to rename columns;

These functions can be combined in a single statement making use of the pipe operator. 

Here are some examples on the use of these verbs applied to the built-in dataset `starwars` that is part of the `tidyverse` package.

Calculate the mean `birth_year` for all the characters:

```{r}
starwars |> 
  summarize(
    mean_birth_year = mean(birth_year, na.rm = TRUE)
    ) |>
  print()
```

Calculate the average height and the average mass by gender:

```{r}
starwars |>
  group_by(gender) |>
  summarize(
    mean_height = mean(height, na.rm = T),
    mean_mass = mean(mass, na.rm = T)
  ) 
```

Now, let's use the verbs to create a graph. We want to create a scatterplot with mass on the x-axis and height on the y-axis. Also, we want to add a third variable in the mix: whether the character's birth year is hither than 50 or not (just for fun). Finally we only want to show characters that have a mass lower than 200 and with a birth_year that is known.

```{r}
starwars |>
  # First filter the cases with a mass higher than 200 and for which birth_year is not NA
  filter(
    mass < 200,
    !is.na(birth_year)
  ) |>

  # Then we create a variable that has value yes if the character's birth year is above 50 and no otherwise
  mutate(
    birth_above_50 = if_else(birth_year > 50, "yes", "no") 
  ) |>
  
  # Select the necessary variables for the graph (not really necessary but this is to show this option)
  select(
    height, 
    mass, 
    birth_above_50) |>

  # Create the graph
  ggplot(
    aes(x = mass, y = height, color = birth_above_50)
  ) + geom_point()
```

# Tuesday

## Column 1

### Outline

In the first part, we will introduce the basic rationale behind Bayesian statistics. We start by explaining the three key components of a Bayesian model: the prior, the likelihood, and the posterior. 

Then, we switch to how the estimation of parameters by first introducing the basic idea of grid approximation and then outlining the basic idea of MCMC sampling.

At the end of the session we introduce `brms` and learn how to estimate a simple regression model with `brms` and just use the `summary()` and `plot()` functions to get insight in the model results.

### Materials

#### Slides

The htlm-version of the slides for this first part can be found <a href="Presentations/Part1/Slides_Part1.html" target="blank">here</a>

#### Data

For this first part, we used a straightforward dataset on predicting racetimes for a marathon. The data can be downloaded <a href="Data/MarathonData.RData" target="blank"> here </a> (right-click to save as). 

## Column 2

### References and resources

Data comes from <a href="https://www.kaggle.com/datasets/girardi69/marathon-time-predictions?resource=download" target="blank">Kaggle</a>

Paul Bürkner’s presentation available on YouTube:<a href="https://www.youtube.com/watch?v=FRs1iribZME" target="blank"> click here</a>

Interactive tool demonstrating MCMC sampling: 
<a href="https://chi-feng.github.io/mcmc-demo/app.html#HamiltonianMC,standard" target="blank"> click here </a>

`brms` homepage: <a href="https://paul-buerkner.github.io/brms/" target="blank"> click here </a>


# Wedsnesday

## Column 1

### Outline

## Column 2

### References and resources

# Thursday

## Column 1

### Outline

## Column 2

### References and resources

# Friday

## Column 1

### Outline

## Column 2

### References and resources

# WAMBS Template

## Column 1

### Info

During the course we introduced the WAMBS (when to Worry and how to Avoid the Misuse of Bayesian Statistics) checklist as a tool to structure your thinking and checking of your Bayesian model(s). 

To see an example of this WAMBS checklist applied to the final model on the Marathon Data click <a href="WAMBS_template/WAMBS_workflow_MarathonData.html" target="blank">here</a>.

The Quarto document to make this report can be found <a href="WAMBS_template/WAMBS_workflow_MarathonData.qmd" target="blank">here</a> (right-click to save).


## Column 2

### References and resources

Depaoli, S., & Van de Schoot, R. (2017). Improving transparency and replication in Bayesian statistics: The WAMBS-Checklist. Psychological methods, 22(2), 240.

Van de Schoot, R., Veen, D., Smeets, L., Winter, S. D., & Depaoli, S. (2020). A tutorial on using the WAMBS checklist to avoid the misuse of Bayesian statistics. Small Sample Size Solutions: A Guide for Applied Researchers and Practitioners; van de Schoot, R., Miocevic, M., Eds, 30-49.


